Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
acknowledgements
Update manuscript.Rmd and manuscript.md

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date:  ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland; ^2^ Department of Biology, Memorial University of Newfoundland, Canada.
output: 
    #word_document
    github_document
---

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators (Bergerud 1985) and access islands during calving
(Bergerud et al. 1990). Despite a large number of coastal and island caribou
herds, limited evidence exists suggesting that caribou swim in the ocean as
readily as they swim in freshwater (see Table 1). A key exception is Peary
caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
Miller 2002). Swimming for quadrupedal terrestrial mammals can be thought of as
a set of trade-offs. Costs include high energy expenditure and risk of drowning.
Specifically, even though caribou are adept swimmers, the energetic expenditure
associated with swimming for quadrupedal mammals is significantly higher than
walking or running (Fish 1993). Additionally, drowning is also possible (Miller
& Gunn 1985) although the proportion of caribou that die from drowning is
unknown. Swimming between islands or from the mainland to nearby islands can be
beneficial. Among the most important benefits are predator avoidance and seeking
under-foraged areas.

We describe visual observation of a single swimming event by a caribou in the
North Atlantic Ocean. This observation led us to investigate the prevalence of
swimming of GPS radio-collared caribou in the Fogo Island archipelago in
Newfoundland, Canada. We posit two potential mechanisms explaining the
trade-offs associated with swimming in the North Atlantic Ocean: forage
limitation and the associated density-dependent habitat selection and predator
avoidance. For caribou living on small islands in the ocean, the same notions
for why caribou swim in freshwater may also explain why caribou swim in the
ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007) have been proposed as potential
explanations.

We observed one swimming event in the field as well as 86 remotely sensed
swimming events from three individual caribou over a three-year period. In
general, caribou swam more frequently to larger islands. Fogo, Perry, and
Eastern Indian islands are the largest, and the most commonly occupied. While
caribou often swim across streams, rivers, and lakes (Leblond et al. 2016), few
observations exist of caribou swimming in the ocean (see Table 1). Of the
existing or inferred observations of caribou swimming in the ocean, none appear
to be as prolific as the number and frequency of swimming events we observed. We
surmise that for caribou living on small islands, the primary drivers of
swimming are likely related to forage limitation and predator avoidance.

@@

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by a National
Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
Discovery Grant to EVW.

Land acknowledgement of Beothuk & Mi'kmaq people?

## References

>>>>

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: |
  | ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
  | ^2^Department of Biology, Memorial University of Newfoundland, Canada
  | \*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
    github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
  #bookdown::word_document2:
  #  reference_docx: template.docx
#bibliography: ../references/knitcite.bib
#csl: csl.csl
---

\newpage

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators (Bergerud 1985) and access islands during calving
(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
thought of as a set of trade-offs. Costs include high energy expenditure and
risk of drowning. Specifically, even though caribou are adept swimmers, the
energetic expenditure associated with swimming for quadrupedal mammals is
significantly higher than walking or running (Fish 1993). Additionally, drowning
is also possible (Miller & Gunn 1985) although the proportion of caribou that
die from drowning is unknown. Despite a large number of coastal and island
caribou herds, limited evidence exists suggesting that caribou swim in the ocean
as readily as they swim in freshwater (see Table 1). A key exception is Peary
caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
Miller 2002).

Here, we describe the visual observation of an adult male caribou swimming
between islands in the Fogo Island archipleago in the North Atlantic Ocean. This
observation prompted us to query a large database of GPS radio-collared caribou
in the Fogo Island caribou herd in Newfoundland, Canada. We identified  86
remotely sensed swimming events between XX islands from three individual caribou
over a three-year period.

We posit two potential mechanisms explaining the
trade-offs associated with swimming in the North Atlantic Ocean: forage
limitation and the associated density-dependent habitat selection and predator
avoidance. For caribou living on small islands in the ocean, the same notions
for why caribou swim in freshwater may also explain why caribou swim in the
ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007) have been proposed as potential
explanations.

While caribou often swim across streams, rivers, and lakes (Leblond et al.
2016), few observations exist of caribou swimming in the ocean (see Table 1). Of
the existing or inferred observations of caribou swimming in the ocean, none
appear to be as prolific as the number and frequency of swimming events we
observed. We surmise that for caribou living on small islands, the primary
drivers of swimming are likely related to forage limitation and predator
avoidance.

@@

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by a National
Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
Discovery Grant to EVW. We respectfully acknowledge the territory in which data
were collected and analyzed as the ancestral homelands of the Beothuk, and the
island of Newfoundland as the ancestral homelands of the Mi’kmaq and Beothuk.

## Literature Cited
